---
title: "Using custom Python versions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using custom Python versions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette describes the three different ways to initialize the R session's connection to Python using the **rMIDAS** package as well as [**reticulate**](https://github.com/rstudio/reticulate).

## Option 1: Do nothing!
 
rMIDAS relies on Python 3.X to run the MIDAS imputation algorithm (note: Python 3.9 is as yet untested for full rMIDAS functionality). For most users, the default settings in **rMIDAS** will be sufficient. Both `train()` and `complete()` check if Python has been initialized and, if not, run the required setup using the best Python 3 version available on your system (as determined by **reticulate**). The first time you run **rMIDAS** after installation, you may be prompted to install additional Python dependencies.

If a suitable Python version is not found on your system, you will be asked to manually set the path to a Python binary. You can do this using the next option.

## Option 2: 

If the automatic setup returns an error or you wish to use a specific Python binary on your system, you can use the `set_python_env()` function in **rMIDAS**, providing an exact path to your chosen Python binary:

```{r, eval = FALSE}

library(rMIDAS)

set_python_env(x = "~/path/to/bin/python")

# Then proceed as normal...

```

With `set_python_env()` you can also set a virtualenv or condaenv environment:

```{r, eval = FALSE}

library(rMIDAS)

set_python_env(x = "myenv", type = "virtualenv")

# or

set_python_env(x = "mycondaenv", type = "conda")

# Then proceed as normal...

```

On the [GitHub repository](https://github.com/MIDASverse/rMIDAS/) you can also find an environment file (`rmidas-env.yml`) which can be used to initialise a new conda environment that contains Python 3.7 and all required dependencies. 

Once you have downloaded this file, in your console navigate to the download directory and run:
```{bash, eval=FALSE}
conda env create -f rmidas-env.yml
```

Then, prior to training a MIDAS model, make sure to load this environment in R:
```{r, eval=FALSE}
set_python_env(x = "rmidas-env", type = "conda")
```

*Note*: **reticulate** only allows you to set a Python binary once per R session, so if you wish to switch to a different Python binary, or have already run `train()` or `convert()`, you will need to restart R prior to using `set_python_env()`.

## Option 3:

If you desire more granular control of the R-Python interface, it is possible to use **reticulate**'s in-built Python configuration tools. Since these commands are outside of **rMIDAS**, you must also manually call `midas_setup()` after configuring your Python install, e.g.:

```{r, eval = FALSE}

library(rMIDAS)

reticulate::use_condaenv(condaenv = "myenv", conda = "some_conda_executable", required = FALSE)
midas_setup()

# Then proceed as normal...

```

As with option 2, **reticulate** only allows you to set a Python binary once per R session. If you wish to switch to a different Python binary, or have already run `train()` or `convert()`, you will need to restart R prior to changing Python version and then call `midas_setup()`.

## Troubleshooting errors

Sometimes the above three options may fail due to system configuration issues. Here we note a few common issues and fixes.

### Mac defaults to Python 2.7

If you are using a Mac, **reticulate** may be defaulting to Python 2.7 which is not supported by **rMIDAS**. If this is the case you will have to configure the R session to use a Python 3 binary, as in option 2 above, by running:

```{r, eval = FALSE}

set_python_env(x = "/usr/local/bin/python3")

# Then proceed as normal...
```

If this returns an error, it's likely **reticulate** cannot find a Python environment related to the binary. In which case we recommend restarting the R session and creating a `virtualenv` that points to your desired Python 3 binary, as follows:

```{r, eval = FALSE}

reticulate::virtualenv_create(envname = "myenv", python = "/path/to/your/python3/bin")
set_python_env(x = "myenv", type = "virtualenv")

# Then proceed as normal...

```

### Shared library access

If, after setting a Python binary/virtualenv/conda environment using either **rMIDAS** or **reticulate**, you still get an error, it is worth calling `reticulate::py_discover_config` to check whether the required python binary is visible.

If the `python` path is correct, but `libpython` is listed as `[NOT FOUND]` this suggests your Python binary does not have a shared library. In which case, either point to an alternative Python binary or reinstall your Python version with shared library enabled. On a Unix/Linux system, and using `pyenv`, this can be done as follows (replacing the version number as required):

```{bash, eval = FALSE}
env PYTHON_CONFIGURE_OPTS="--enable-shared" pyenv install 3.8.6
```
