---
title: "Using custom Python versions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using custom Python versions}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This vignette describes the three different ways to initialize the R session's connection to Python using the **rMIDAS** package as well as [**reticulate**](https://github.com/rstudio/reticulate).

## Option 1: Do nothing!
 
rMIDAS relies on Python to run the MIDAS imputation algorithm. Currently, Python versions from 3.6 to 3.10 are supported. For most users, the default settings in **rMIDAS** will be sufficient. Both `train()` and `complete()` check if Python has been initialized. The first time you call **rMIDAS** after installation, you will be prompted on whether to set up the Python environment and dependencies automatically.

If the automatic setup fails or you want to set up your Python environment manually, you can do this using the next option.

*Note*: If you are using **rMIDAS** in headless mode (for example using Amazon AWS), the automatic setup will not execute. In this instance you need to set up the Python environment and its dependencies manually. Please refer to our *Running rMIDAS on a server instance* vignette for additional help.

## Option 2: 

If the automatic setup returns an error or you wish to use a specific Python binary on your system, you can use the `set_python_env()` function in **rMIDAS**, providing an exact path to your chosen Python binary:

```{r, eval = FALSE}

library(rMIDAS)

# Decline the automatic setup

set_python_env(x = "~/path/to/bin/python")

# Then proceed as normal...

```

With `set_python_env()` you can also set a virtualenv or conda environment:

```{r, eval = FALSE}

library(rMIDAS)

# Decline the automatic setup

set_python_env(x = "myenv", type = "virtualenv")

# or

set_python_env(x = "mycondaenv", type = "conda")

# Then proceed as normal...

```

On the [rMIDAS GitHub repository](https://github.com/MIDASverse/rMIDAS/) you can also find an environment file (`rmidas-env.yml`) which can be used to initialise a new conda environment that contains Python 3.8 and all required dependencies. 

Once you have downloaded this file, in your console navigate to the download directory and run:
```{bash, eval=FALSE}
conda env create -f rmidas-env.yml
```

Then, prior to training a MIDAS model, make sure to load this environment in R:
```{r, eval=FALSE}
set_python_env(x = "rmidas", type = "conda")
```

*Note*: **reticulate** only allows you to set a Python binary once per R session, so if you wish to switch to a different Python binary, or have already run `train()` or `convert()`, you will need to restart or terminate R prior to using `set_python_env()`.

## Option 3:

If you desire more granular control of the R-Python interface, it is possible to use **reticulate**'s in-built Python configuration tools. Since these commands are outside of **rMIDAS**, you must also manually call `midas_setup()` after configuring your Python install, e.g.:

```{r, eval = FALSE}

library(rMIDAS)

# Decline the automatic setup

reticulate::use_condaenv(condaenv = "myenv", conda = "some_conda_executable", required = FALSE)
midas_setup()

# Then proceed as normal...

```

As with Option 2, **reticulate** only allows you to set a Python binary once per R session. If you wish to switch to a different Python binary, or have already run `train()` or `convert()`, you will need to restart R prior to changing Python version and then call `midas_setup()`.

## Troubleshooting errors

Sometimes the above three options may fail due to system configuration issues. Here we note a few common issues and fixes.

### Older versions of macOS default to Python 2.7

Macs used to come with Python 2.7 pre-installed between macOS versions 10.8 and 12.3. If you are using a Mac, **reticulate** may be defaulting to Python 2.7 which is not supported by **rMIDAS**. If this is the case you will have to configure the R session to use a Python 3 binary, as in option 2 above, by running:

```{r, eval = FALSE}

set_python_env(x = "/usr/local/bin/python3")

# Then proceed as normal...
```

If this returns an error, it's likely **reticulate** cannot find a Python environment related to the binary. In which case we recommend restarting the R session and creating a `virtualenv` that points to your desired Python 3 binary, as follows:

```{r, eval = FALSE}

reticulate::virtualenv_create(envname = "myenv", python = "/path/to/your/python3/bin")
set_python_env(x = "myenv", type = "virtualenv")

# Then proceed as normal...

```

### Shared library access

If, after setting a Python binary/virtualenv/conda environment using either **rMIDAS** or **reticulate**, you still get an error, it is worth calling `reticulate::py_discover_config` to check whether the required python binary is visible.

If the `python` path is correct, but `libpython` is listed as `[NOT FOUND]` this suggests your Python binary does not have a shared library. In which case, either point to an alternative Python binary or reinstall your Python version with shared library enabled. On a Unix/Linux system, and using `pyenv`, this can be done as follows (replacing the version number as required):

```{bash, eval = FALSE}
env PYTHON_CONFIGURE_OPTS="--enable-shared" pyenv install 3.8.6
```
